---
title: "Project Bellabeat"
author: "Juan Pablo Montilla"
date: "2025-07-11"
output:
  pdf_document:
    toc: true
    number_sections: true
    toc_depth: 2
    highlight: tango
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---


## 1. Introduction
For this project, the stakeholder at Bellabeat wants to analyze trends in health-focused smart devices. Using this analysis, we aim to understand how the insights derived from Fitbit data can be applied to Bellabeat's product portfolio and how they may help shape Bellabeat's future marketing strategies.
## 2. Data Loading


We are using a public dataset: Fitbit Fitness Tracker Data (CC0: Public Domain, made available via Mobius). This dataset contains minute-level data on physical activity, sleep, calories and heart rate. It was used in the Google Data Analytics Capstone as a proxy to analyze user behavior for Bellabeat.

The files were downloaded, unzipped and loaded into R using `read_csv()`:

```{r load-data}
library(readr)

dailyActivity_merged <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
daily_Calories <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv")
daily_Intensities <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv")
daily_Steps <- read_csv("Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv")
hourly_Calories <- read_csv("Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")
minute_Sleep <- read_csv("Fitabase Data 4.12.16-5.12.16/minuteSleep_merged.csv")
```

## 3. Cleaning and exploration
in this section, I cleaned and explored the `dailyActivity_merged` dataset, which includes steps, calories, distances and minutes of activity.

```{r load-libraries}
library(tidyverse)
library(lubridate)
```
## 3.1 Data Preparation

# Load dataset
```{r read}
daily <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
```
# Fix column names and format date
```{r format}
daily <- daily %>%
  mutate(ActivityDate = as.Date(ActivityDate, format = "%m/%d/%Y")) %>%
  rename(
    user_id = Id,
    date = ActivityDate,
    total_steps = TotalSteps,
    total_distance = TotalDistance,
    tracker_distance = TrackerDistance,
    logged_active_distance = LoggedActivitiesDistance,
    very_active_minutes = VeryActiveMinutes,
    fairly_active_minutes = FairlyActiveMinutes,
    lightly_active_minutes = LightlyActiveMinutes,
    sedentary_minutes = SedentaryMinutes,
    calories = Calories
  )
```
## 3.2 Summary statistics
```{r summary}
summary(daily)
```
## 4. Behavior Analysis
Now I want to understand the behavior of each user in terms of activity and calories burned. This helps identify different user types and how Bellabeat could tailor its marketing accordingly.
```{r behavior}
user_summary <- daily %>%
  group_by(user_id) %>%
  summarise(
    days_tracked = n(),
    avg_steps = mean(total_steps),
    avg_calories = mean(calories),
    avg_sedentary_min = mean(sedentary_minutes),
    avg_active_min = mean(very_active_minutes + fairly_active_minutes + lightly_active_minutes)
  )
```
## 4.1 Histograms: Step distribution
```{r steps}
ggplot(user_summary, aes(x = avg_steps)) +
  geom_histogram(binwidth = 1000, fill = "#69b3a2", color = "white") +
  labs(title = "Average Daily Steps per User", x = "Steps", y = "Users") +
  theme_minimal()
```
## 4.2 Step VS Calories

```{r comparison}
ggplot(user_summary, aes(x = avg_steps, y = avg_calories)) +
  geom_point(alpha = 0.6, color = "#1f77b4") +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red") +
  labs(title = "Steps vs. Calories", x = "Average Steps", y = "Average Calories") +
  theme_minimal()
```
## 5. User segmentation
To improve the marketing strategy, I created four activity segments based on users' average daily steps.

```{r segmentation}
user_summary <- user_summary %>%
  mutate(activity_level = case_when(
    avg_steps < 5000 ~ "Low",
    avg_steps < 10000 ~ "Moderate",
    avg_steps < 15000 ~ "Active",
    TRUE ~ "Very Active"
  ))
```
## 5.1 Bar chart: Activity levels
```{r actitities}
ggplot(user_summary, aes(x = activity_level)) +
  geom_bar(fill = "#4e79a7") +
  labs(title = "Users by Activity Level", x = "Activity Level", y = "Number of Users") +
  theme_minimal()
```

## 6. Insights and recommendations
 - Over 70% of users record less than 10,000 steps per day → opportunity for motivational marketing campaigns

- Clear correlation between steps and calories burned → can guide personalized feedback

- Sedentary time is high even in active users → opportunity to promote micro-movements or stretch reminders

- Bellabeat can use this segmentation to:

    -Design custom challenges per user group

    -Increase product engagement via targeted in-app notifications

    -Tailor its wellness content and product design to real user behavior
    
## 7. Author
Juan Pablo Montilla Montaña
Portafolio: 
